---
title: "Untitled"
output: html_document
date: "2024-05-04"
---

```{r}
library(dplyr)
library(ggplot2)
```

GWAS QC:

Identification of individuals with missing data rates or heterozygosity rate
```{r}
missing <- read.table("/Users/albacoto/Desktop/finalproject/GWAS-QC1.imiss",header=T)

# N_MISS: number of missing SNPs 
# F_MISS: proportion of missing SNPs per individual
```

```{r}
het <- read.table("/Users/albacoto/Desktop/finalproject/GWAS-QC1.het",header=T)

# O.HOM: homozygous genotypes 
# N(NM): non-missing genotypes per individual
```

Calculate heterozygosity rate
```{r}
#Het = (N(NM) − O(Hom))/N(NM)
het$het_rate = (het$N.NM.-het$O.HOM.)/het$N.NM.
```

```{r}
QC1 <- inner_join(missing,het)
```

Visualization:
```{r}
ggplot(QC1, aes(x = het_rate, y = F_MISS)) +
  geom_point() +
  labs(x = "Observed Heterozygosity Rate",
       y = "Proportion of Missing SNPs",
       title = "Observed Heterozygosity vs. Proportion of Missing SNPs")
```

```{r}
# mean and sd to see in the plot heterozygosity rate that is more than 3 s.d. from the mean
mean_het = mean(QC1$het_rate, na.rm = TRUE)
sd_het = sd(QC1$het_rate, na.rm = TRUE)
print(mean_het)
print(sd_het)
```

```{r}
ggplot(QC1, aes(x = het_rate, y = F_MISS)) +
  geom_point(color = "blue", alpha = 0.6, size = 1.5) +  # Add transparency and adjust size
  labs(
    x = "Observed Heterozygosity Rate",
    y = "Proportion of Missing SNPs",
    title = "Observed Heterozygosity vs. Proportion of Missing SNPs"
  ) +
  geom_hline(yintercept = 0.97, linetype = "dashed", color = "red", size = 1) + # Adjust line type and size
  geom_vline(xintercept = mean_het + 3 * sd_het, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = mean_het - 3 * sd_het, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = mean_het + 3 * sd_het, y = max(QC1$F_MISS), label = "+3 SD", hjust = -0.1, color = "red") + # Add labels for lines
  annotate("text", x = mean_het - 3 * sd_het, y = max(QC1$F_MISS), label = "-3 SD", hjust = 1.1, color = "red") +
  annotate("text", x = min(QC1$het_rate), y = 0.03, label = "0.03 Threshold", vjust = -1, color = "red") +
  theme_minimal() +  # Use a minimal theme for better aesthetics
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```



```{r}
# Remove individuals from the dataset
right_area = mean_het + 3*sd_het
left_area =  mean_het - 3*sd_het
```

```{r}
outliers = subset(QC1, QC1$het_rate > right_area | QC1$het_rate < left_area | QC1$F_MISS >= 0.97)

# make a file to later remove FID & IID columns 
write.table(outliers[,c(1,2)], file = "/Users/albacoto/Desktop/finalproject/wrong_het_missing.txt", col.names = F, row.names = F)
```


Identification of duplicated or related individuals
```{r}
ibd <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC2.genome', header = TRUE)
members <- ibd$FID1
members <- unique(members)
```

```{r}
write.table(cbind(members,members), file = '/Users/albacoto/Desktop/finalproject/wrong_ibd.txt', col.names = F, row.names = F)
```


Visualization
```{r}
# Scatter plot of PI_HAT values
ggplot(ibd, aes(x = PI_HAT, y = FID1)) +
  geom_point(alpha = 0.7, color = "blue") +
  labs(
    title = "PI_HAT Values for Individual",
    x = "PI_HAT Value",
    y = "Individual (FID1)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r}
# Plot the PI_HAT values
ggplot(ibd, aes(x = PI_HAT)) +
  geom_histogram(binwidth = 0.01, fill = "blue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of PI_HAT Values",
    x = "PI_HAT Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

```{r}
# Filter rows where PI_HAT is greater than or equal to 0.5 and select the first two columns
ibd_filtered <- ibd %>%
  filter(PI_HAT >= 0.5) %>%
  select(1, 2) # Assuming the first two columns are the ones you need
```

```{r}
# Write the filtered data to a text file
write.table(ibd_filtered, "/Users/albacoto/Desktop/finalproject/ibd_filtered.txt", row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t")
```



SNPs with an excessive missing data rate
```{r}
missing_SNP = read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC3.lmiss', header = T)
```

```{r}
# Create the histogram 
ggplot(missing_SNP, aes(x = F_MISS)) +
  geom_histogram(binwidth = 0.01, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of SNP Missing Data Rates",
    x = "Missing Data Rate (F_MISS)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  geom_density(aes(y = ..count.. * 0.01), color = "red", size = 1, adjust = 1) +
  theme(
    panel.grid.major = element_line(color = "gray", linetype = "dashed"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  )
```


PCA
```{r}
pca <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC2.eigenvec', head = T)
names(pca) <- c(c("FID", "IID"), paste(rep("PC", 20), seq(1,20), sep=""))
```

```{r}
# Extract PC1 and PC2 from the 'pca' dataset
PC1 <- pca$PC1
PC2 <- pca$PC2

# Create a scatter plot
plot(PC1, PC2,
     xlab = "PC1",
     ylab = "PC2",
     main = "Principal Component Analysis",
     col = "skyblue")   
```

```{r}
# Create the plot using ggplot2
ggplot(pca, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7, size = 2) +  # Adjust alpha and size for better visibility
  labs(
    x = "Principal Component 1 (PC1)",
    y = "Principal Component 2 (PC2)",
    title = "PCA Plot: PC1 vs PC2"
  ) +
  theme_minimal(base_size = 15) +  # Use a minimal theme with a larger base font size
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +  # Add a dashed horizontal line at y = 0
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey")  # Add a dashed vertical line at x = 0

```

Variance with eigenvalues
```{r}
eigenvalues <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC2.eigenval', head = T)
```


```{r}
# Compute the total variance
total_variance <- sum(eigenvalues)

# Compute the variance explained by each PC
variance_explained <- eigenvalues / total_variance
percentage_variance_explained <- variance_explained * 100

# Percentage of variance explained by the first PC
percentage_variance_explained[1]

```

```{r}
pca_outlier <- pca %>% filter(PC1 > 0.95)

#outliers saved for removal
write.table(pca_outlier[,c(1,2)], 'pca_outlier.txt', col.names = FALSE, row.names = FALSE)
```


EYE COLOR
```{r}
eye_color <- read.table("/Users/albacoto/Desktop/finalproject/eye_color.txt", header=F)
colnames(eye_color) <- c("IID", "Eye color")
eye_color$pheno <- ifelse(grepl("brown|black", eye_color$'Eye color', ignore.case = TRUE), 1, 2)

eye_color_id <- eye_color %>% mutate(FID = IID) %>% select("IID", "FID", "pheno")
write.table(eye_color_id, 'eye_color_id.txt', col.names = FALSE, row.names = FALSE)
```


```{r}
eye_color_summary <- as.data.frame(table(eye_color$"Eye color"))
colnames(eye_color_summary) <- c("Eye color", "Count")
eye_color_summary
```


Create a binary phenotype:
```{r}
# Get unique values in the "eye color" column
unique_eye_colors <- unique(eye_color$"Eye color")
print(unique_eye_colors)

eye_color_counts <- table(eye_color$"Eye color")

# Convert eye color counts to data frame for plotting
eye_color_df <- data.frame(Eye_Color = names(eye_color_counts),
                           Frequency = as.numeric(eye_color_counts))

# Plot the bar plot with rotated x-axis labels and adjusted y-axis limits
barplot(eye_color_df$Frequency, 
        names.arg = eye_color_df$Eye_Color,
        ylab = "Frequency",
        main = "Distribution of Eye Colors",
        col = "skyblue",        # Change color as needed
        las = 2,                # Rotate x-axis labels by 90 degrees
        ylim = c(0, 500),       # Set y-axis limits from 0 to 500
        cex.names = 0.5)        # Adjust font size of axis labels if necessary

```


ADD THE PHENOTYPES
```{r}
phenotypes <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC4.fam', header = F)
names(phenotypes) <- c("FID", "IID", "ID_father", "ID_mother", "sex", "pheno")

PCA2 <- inner_join(phenotypes, pca)
```

```{r}
#add pheno
PCA3 <- merge(eye_color_id, PCA2, by = "IID", all.x = TRUE)
```


```{r}
# Identify rows with NA values
na_rows <- PCA3[!complete.cases(PCA3), ]

# Save the IIDs of the rows with NA values to a text file
write.table(na_rows[, c("FID.x", "IID")], "na_iids.txt", col.names = FALSE, row.names = FALSE, quote = FALSE)
```


```{r}
# remove NA rows
PCA3 <- PCA3 %>% filter(!is.na(PCA3$'FID.y'))
```


more PCA
```{r}
pca_filtered <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC5.eigenvec', head = T)
names(pca_filtered) <- c(c("FID", "IID"), paste(rep("PC", 20), seq(1,20), sep=""))
```

```{r}
# Select the specific columns from PCA3
PCA3_selected <- PCA3 %>% select(IID, pheno.x, ID_father, ID_mother, sex)

# Perform the inner join with pca_filtered on the common column FID
pca_filtered2 <- inner_join(pca_filtered, PCA3_selected, by = "IID")
```

```{r}
# Enhanced plot
ggplot(pca_filtered2, aes(x = PC1, y = PC2, color = as.factor(pheno.x))) + 
  geom_point(size = 3, alpha = 0.6) +  # Increase point size and set transparency
  labs(x = "PC1", y = "PC2", color = "Eye Color") +  # Add axis labels
  scale_color_manual(values = c("brown", "skyblue"), labels = c("Brown", "Blue")) +  # Define color palette and labels
  theme_minimal() +  # Use a minimal theme for cleaner appearance
  theme(
    legend.position = "bottom",  # Position the legend at the bottom
    legend.title = element_text(size = 12, face = "bold"),  # Adjust legend title appearance
    legend.text = element_text(size = 10),  # Adjust legend text appearance
    axis.text = element_text(size = 10),  # Adjust axis text size
    axis.title = element_text(size = 12, face = "bold")  # Adjust axis title size and style
  )

```

Variance
```{r}
eigenvalues2 <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC5.eigenval', head = F)
names(eigenvalues2) <- c("Eigenvalues")
```

```{r}
#variance explained by every PC
variance_explained2 <- (eigenvalues2$Eigenvalues / sum(eigenvalues2$Eigenvalues)) * 100 
```


```{r}
variance_explained2 <- data.frame(PC = seq_along(variance_explained2), Variance = variance_explained2)

# plot of variance explained by PCs
ggplot(variance_explained2, aes(x = PC, y = Variance)) +
  geom_line(color = "blue", size = 1) +  # Set line color and size
  geom_point(color = "red", size = 2) +  # Set point color and size
  labs(
    x = "Principal Component (PC)",
    y = "Percentage of Variance Explained",
    title = "Percentage of Variance Explained by Principal Component"
  ) + 
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center and style the title
    axis.title.x = element_text(size = 14, face = "bold"),  # Style the x-axis title
    axis.title.y = element_text(size = 14, face = "bold"),  # Style the y-axis title
    axis.text = element_text(size = 12)  # Adjust axis text size
  ) 

```


QC of SNPs
```{r}
missing2 = read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC5.lmiss', header = T)
```


```{r}
ggplot(missing2, aes(x = F_MISS)) +
  geom_histogram(binwidth = 0.01, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of SNP Missing Data Rates",
    x = "Missing Data Rate (F_MISS)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  geom_density(aes(y = ..count.. * 0.01), color = "red", size = 1, adjust = 1) +
  theme(
    panel.grid.major = element_line(color = "gray", linetype = "dashed"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  )
```

```{r}
missing2 %>% filter(F_MISS >= 0.05) %>% nrow()
```

```{r}
test_missing = read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC6.missing', header = TRUE)
```

```{r}
fail_diffmiss_qc = test_missing[test_missing$P < 10e-5, 2]
fail_diffmiss_qc 
write.table(fail_diffmiss_qc, file = 'fail-diffmiss-qc.txt', row.names = F, col.names = F)
```


GWAS ASSOCIATION
```{r}
fisher <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC7.assoc.fisher', head = T)
```


```{r}
#p-value and location of the most significant variant
fisher[which.min(fisher$P),]$SNP
min(fisher$P)
fisher[which.min(fisher$P),]$BP
fisher[which.min(fisher$P),]$CHR
```

```{r}
#bonferroni correction
bon <- 0.00001/nrow(fisher)
bon
#number of tests
nrow(fisher)
#test if any SNPs significant after bonferroni
min(fisher$P) <= bon
```


```{r}
library(qqman)
manhattan(fisher)
```

```{r}
qq(fisher$P)
```


GENOMIC CONTROL:
```{r}
fisher$ChiSq <- qchisq(fisher$P, df = 1, lower.tail = FALSE)
```

```{r}
# INFLATION FACTOR
lambda <- median(fisher$ChiSq) / qchisq(0.5, df = 1)
print(lambda)  
```

```{r}
#adjust chi-squared
fisher$Adjusted_ChiSq <- fisher$ChiSq / lambda

# convert chi back to p-values
fisher$Adjusted_P <- pchisq(fisher$Adjusted_ChiSq, df = 1, lower.tail = FALSE)
```

```{r}
# adjusted p-value of most significant marker
most_significant_marker <- fisher[fisher$SNP == "rs4778241", ]
adjusted_p_value <- most_significant_marker$Adjusted_P
print(adjusted_p_value)
```



ADJUSTED
```{r}
fisher_adj <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC7.assoc.fisher.adjusted', head = T)

fisher_adj_joined <- inner_join(fisher, fisher_adj)
```


```{r}
# bonferri correction significance:
min(fisher_adj$BONF)
```


```{r}
# convert p-values to quantiles under Chi 
gc_fish_joined <- fisher_adj_joined %>% 
mutate(q = qchisq(P, df = 1, lower.tail = F))

# medians for Chi
exp_median <- qchisq(0.5, df = 1, lower.tail = F)
exp_median

obs_median <- median(gc_fish_joined$q)
obs_median

# Inflation factor 
lamda <- obs_median/exp_median
lamda
```



LOGISTIC
```{r}
pca_adj <- read.table('/Users/albacoto/Desktop/finalproject/adjust1pc.assoc.logistic', head = T)

# p-values for SNPs (ADD) and for covariates (COV1); only use ADD
pca_adj <- pca_adj %>% filter(TEST == 'ADD')
pca_adj <- pca_adj %>%
  filter(!is.na(P) & is.finite(P))
```

```{r}
manhattan(pca_adj)
```

```{r}
# Sort the data frame by p-value
sorted_pca_adj <- pca_adj[order(pca_adj$P), ]

# Get the top SNPs
top_snps <- head(sorted_pca_adj, n = 20) 
```


```{r}
#qq after PC adjustment
qq(pca_adj$P)
```

```{r}
#significant SNPs under bonferroni p-value treshold
sign <- pca_adj %>% filter(pca_adj$P < bon)
sign
sign[which.min(sign$P),]$SNP
```

```{r}
manhattan(subset(pca_adj, CHR == 15), highlight = sign$SNP, xlim = c(min(sign$BP) - 10000,
    max(sign$BP) + 10000), main = "Chr 15", ylim = c(0,50))
```

```{r}
#adjusted inflation factor, lamda value
pca_adj <- pca_adj %>% 
mutate(q = qchisq(P, df = 1, lower.tail = F))

exp_median <- qchisq(0.5, df = 1, lower.tail = F)

lamda2 <- median(pca_adj$q)/exp_median
lamda2
```



FURTHER ANALYSIS:
```{r}
# Load the conditional association results
fisher_conditioned <- read.table('/Users/albacoto/Desktop/finalproject/GWAS-QC7_condition.assoc.logistic', header = TRUE)
```


```{r}
fisher_conditioned <- fisher_conditioned %>% filter(TEST == 'ADD')

fisher_conditioned <- fisher_conditioned %>%
  filter(!is.na(P) & is.finite(P))
```
```{r}
manhattan(fisher_conditioned, ylim = c(0,21))
```


```{r}
# QQ plot for the conditioned analysis
qq(fisher_conditioned$P)
```


```{r}
# Identify the new most significant SNP
new_most_significant_snp <- fisher_conditioned[which.min(fisher_conditioned$P), ]
print(new_most_significant_snp)

```

REGION OF MOST SIGNIFICANT SNP
```{r}
library(data.table)

genotype_data <- fread("/Users/albacoto/Desktop/finalproject/significant_snp_window.raw")

# Merge the genotype and phenotype data by the sample ID
merged_data <- merge(genotype_data, eye_color_id, by = "IID")

head(merged_data)
```


SNP: rs7179994_G
```{r}
# Summary statistics for the most significant SNP, excluding NAs
summary_stats1 <- merged_data %>%
  filter(!is.na(rs7179994_G) & !is.na(pheno)) %>%
  group_by(rs7179994_G, pheno) %>%
  summarise(count = n()) %>%
  ungroup()
```


```{r}
# Remove NAs for the plot
filtered_data1 <- merged_data %>%
  filter(!is.na(rs7179994_G) & !is.na(pheno))

# Plot
ggplot(filtered_data1, aes(x = as.factor(rs7179994_G), fill = as.factor(pheno))) +
  geom_bar(position = "dodge", color = "black") +  # Adding a black border to bars
  scale_fill_manual(values = c("1" = "brown", "2" = "skyblue"), 
                    labels = c("1" = "Brown", "2" = "Blue")) +  # Custom colors and labels
  labs(title = "Distribution of Eye Color by rs7179994_G Genotype",
       x = "rs7179994_G Genotype",
       y = "Count",
       fill = "Eye Color") +
  theme_minimal(base_size = 15) +  # Increase base font size for better readability
  theme(plot.title = element_text(hjust = 0.5))  # Center align the title

```

SNP: rs7174027_A
```{r}
# Summary statistics for the most significant SNP, excluding NAs
summary_stats2 <- merged_data %>%
  filter(!is.na(rs7174027_A) & !is.na(pheno)) %>%
  group_by(rs7174027_A, pheno) %>%
  summarise(count = n()) %>%
  ungroup()
```


```{r}
# Remove NAs for the plot
filtered_data2 <- merged_data %>%
  filter(!is.na(rs7174027_A) & !is.na(pheno))

# Plot
ggplot(filtered_data2, aes(x = as.factor(rs7174027_A), fill = as.factor(pheno))) +
  geom_bar(position = "dodge", color = "black") +  # Adding a black border to bars
  scale_fill_manual(values = c("1" = "brown", "2" = "skyblue"), 
                    labels = c("1" = "Brown", "2" = "Blue")) +  # Custom colors and labels
  labs(title = "Distribution of Eye Color by rs7174027_A Genotype",
       x = "rs7174027_A Genotype",
       y = "Count",
       fill = "Eye Color") +
  theme_minimal(base_size = 15) +  # Increase base font size for better readability
  theme(plot.title = element_text(hjust = 0.5))  # Center align the title

```

SNP: rs4778241_A
```{r}
# Summary statistics for the most significant SNP, excluding NAs
summary_stats3 <- merged_data %>%
  filter(!is.na(rs4778241_A) & !is.na(pheno)) %>%
  group_by(rs4778241_A, pheno) %>%
  summarise(count = n()) %>%
  ungroup()
```


```{r}
# Remove NAs for the plot
filtered_data3 <- merged_data %>%
  filter(!is.na(rs4778241_A) & !is.na(pheno))

# Plot
ggplot(filtered_data3, aes(x = as.factor(rs4778241_A), fill = as.factor(pheno))) +
  geom_bar(position = "dodge", color = "black") +  # Adding a black border to bars
  scale_fill_manual(values = c("1" = "brown", "2" = "skyblue"), 
                    labels = c("1" = "Brown", "2" = "Blue")) +  # Custom colors and labels
  labs(title = "Distribution of Eye Color by rs4778241_A Genotype",
       x = "rs4778241_A Genotype",
       y = "Count",
       fill = "Eye Color") +
  theme_minimal(base_size = 15) +  # Increase base font size for better readability
  theme(plot.title = element_text(hjust = 0.5))  # Center align the title

```

SNP: rs7495174_G
```{r}
# Summary statistics for the most significant SNP, excluding NAs
summary_stats4 <- merged_data %>%
  filter(!is.na(rs7495174_G) & !is.na(pheno)) %>%
  group_by(rs7495174_G, pheno) %>%
  summarise(count = n()) %>%
  ungroup()
```


```{r}
# Remove NAs for the plot
filtered_data4 <- merged_data %>%
  filter(!is.na(rs7495174_G) & !is.na(pheno))

# Plot
ggplot(filtered_data4, aes(x = as.factor(rs7495174_G), fill = as.factor(pheno))) +
  geom_bar(position = "dodge", color = "black") +  # Adding a black border to bars
  scale_fill_manual(values = c("1" = "brown", "2" = "skyblue"), 
                    labels = c("1" = "Brown", "2" = "Blue")) +  # Custom colors and labels
  labs(title = "Distribution of Eye Color by rs7495174_G Genotype",
       x = "rs7495174_G Genotype",
       y = "Count",
       fill = "Eye Color") +
  theme_minimal(base_size = 15) +  # Increase base font size for better readability
  theme(plot.title = element_text(hjust = 0.5))  # Center align the title

```







